---
title: "Untitled"
format: html
---

```{r}

options(repos = c(CRAN = "https://cran.rstudio.com"))
install.packages("tidyverse")
library(tidyverse)

```

```{r}

df <- read.csv("Data/processed-data/Senators_sentiment.csv")

```

```{r}

senator_summary <- df %>%
  mutate(full_name = paste(first, last)) %>%  
  distinct(state, full_name, party) %>%  
  arrange(state, full_name) %>%             
  group_by(state) %>%
  summarise(
    Senators = paste(paste(full_name, paste0("(", party, ")"), sep = " "), 
                     collapse = ", "),       
    Count = n()               
  ) %>%
  ungroup()
  
print(senator_summary, n = Inf)

```

```{r}

independent_counts <- df %>%
  filter(!is.na(sentiment)) %>%
  count(sentiment)

independent_counts <- independent_counts %>%
  mutate(sentiment = factor(sentiment, levels = c(-1, 1), labels = c("Risk", "Opportunity")))

ggplot(independent_counts, aes(x = sentiment, y = n, fill = sentiment)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("Risk" = "red", "Opportunity" = "#078d0f")) +
  labs(
    title = "Distribution of Sentiments",
    x = "Sentiment",
    y = "Count",
    fill = "Sentiment Type"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

Opportunity sentiment is more prevalent than risk sentiment, indicating that the Senate elected in the 2024 election generally has a proactive stance on addressing challenges and pursuing solutions.

```{r}

grouped_data <- df %>%
  filter(!is.na(sentiment)) %>%
  group_by(party, sentiment) %>%
  summarise(Count = n(), .groups = 'drop') %>%
  group_by(party) %>%
  mutate(Proportion = Count / sum(Count)) 

grouped_data <- grouped_data %>%
  mutate(sentiment = factor(sentiment, levels = c(-1, 1), labels = c("Risk", "Opportunity")))

# Mosaic Plot
ggplot(grouped_data, aes(x = party, y = Proportion, fill = sentiment)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_fill_manual(values = c("Risk" = "red", "Opportunity" = "#078d0f")) +
  labs(
    title = "Proportional Distribution of Sentiments by Party",
    x = "Party",
    y = "Proportion",
    fill = "Sentiment Type"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 0),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5)
  )

```

Democratic and Indepdent senators overwhelmingly view climate change as a very real challenge that they want to tackle. Republicans view addressing climate change as more of a risk. 

```{r}

heatmap_data <- df %>%
  mutate(
    ideology_bin = cut(ideology_score, 
                      breaks = seq(-1, 1, by = 0.2),
                      labels = c("-1.0 to -0.8", 
                               "-0.8 to -0.6", 
                               "-0.6 to -0.4", 
                               "-0.4 to -0.2", 
                               "-0.2 to 0.0",
                               "0.0 to 0.2",
                               "0.2 to 0.4",
                               "0.4 to 0.6",
                               "0.6 to 0.8",
                               "0.8 to 1.0")),
    sentiment = ifelse(sentiment == 1, "Opportunity", "Risk")
  ) %>%
  group_by(ideology_bin, sentiment) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(ideology_bin) %>%
  mutate(percentage = (count / sum(count)) * 100) %>%
  ungroup()

heatmap_data <- heatmap_data %>%
  filter(!is.na(ideology_bin))

# Heatmap
ggplot(heatmap_data, aes(x = sentiment, y = ideology_bin, fill = percentage)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.1f", percentage)), 
            color = "black", size = 4) +
  scale_fill_gradientn(
    colors = c("darkred", "red", "yellow", "lightgreen", "darkgreen"),  
    values = scales::rescale(c(0, 25, 50, 75, 100))  
  ) +
  labs(
    title = "Heatmap of Ideology Score Ranges by Sentiment (Risk and Opportunity)",
    x = "Sentiment",
    y = "Ideology Score Range",
    fill = "Percentage"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    axis.text.y = element_text(hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )

```

The ideology score range represents the political spectrum of the senators, -1.0 to -0.6: Strongly liberal
0.0 to 0.2: Moderately liberal
0.2 to 0.4: Slightly conservative
0.4 to 0.6: Moderately conservative
0.6 to 1.0: Strongly conservative. 
Risk sentiment is not introduced until the 0.2-0.4 range, with sentiment getting more and more risky. It is fully risk from 0.8 to 1. 

```{r}

install.packages("tidyverse")
install.packages("tidyr")
install.packages("reshape2")
install.packages("usmap")
install.packages("dplyr")
library(tidyverse)
library(tidyr)
library(reshape2)
library(usmap)
library(dplyr)

```

```{r}

library(dplyr)
library(usmap)
library(ggplot2)

df <- df %>%
  mutate(state = tolower(state.name[match(state, state.abb)])) 

state_senators <- df %>%
  distinct(state, party) %>%                           
  group_by(state) %>%
  summarise(
    party_composition = case_when(
      all(party == "R") ~ "Republican",
      all(party == "D") ~ "Democrat",
      any(party == "I") & all(party %in% c("I", "D")) ~ "Split (Independent & Democrat)", 
      any(party == "I") & all(party %in% c("I", "R")) ~ "Split (Independent & Republican)",
      TRUE ~ "Split (Republican & Democrat)"
    ),
    .groups = "drop"
  )

party_colors <- c(
  "Republican" = "red",
  "Democrat" = "blue",
  "Split (Independent & Democrat)" = "yellow",
  "Split (Independent & Republican)" = "orange",
  "Split (Republican & Democrat)" = "purple"
)

plot_usmap(data = state_senators, values = "party_composition", regions = "states") +
  scale_fill_manual(
    values = party_colors,
    name = "Party Composition"
  ) +
  labs(
    title = "Senator Party Composition by State",
    subtitle = "Republican, Democrat, Independent, or Split"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    panel.background = element_rect(fill = "white"),
    panel.grid = element_blank()
  )

```

Most states elect senators from the same party, with the exception of Wisconsin, Pennsylvania, Maine, and Vermont.

```{r}

state_sentiment <- df %>%
  filter(!is.na(sentiment)) %>%                
  group_by(state, sentiment) %>%
  summarise(count = n(), .groups = "drop") %>% 
  group_by(state) %>%
  mutate(
    total = sum(count),                        
    proportion = count / total               
  ) %>%
  filter(proportion == max(proportion)) %>%    
  mutate(sentiment_label = case_when(          
    sentiment == -1 ~ "risk",
    sentiment == 1 ~ "opportunity"
  )) %>%
  select(state, sentiment_label)

plot_usmap(data = state_sentiment, values = "sentiment_label", regions = "states") +
  scale_fill_manual(
    values = c("opportunity" = "green", 
               "risk" = "red"),               
    name = "Sentiment",
    breaks = c("opportunity", "risk"),  
    labels = c("Opportunity", "Risk")        
  ) +
  labs(
    title = "Dominant Sentiment by State", 
    subtitle = "Opportunity vs. Risk"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12)
  )

```

States with split Senators have opportunity as their dominant sentiment. 

```{r}

state_sentiment <- df %>%
  filter(!is.na(sentiment)) %>%               
  group_by(state, sentiment) %>%
  summarise(count = n(), .groups = "drop") %>% 
  group_by(state) %>%
  mutate(
    total = sum(count),                        
    proportion = count / total              
  ) %>%
  summarise(
    max_proportion = max(proportion),       
    sentiment_label = case_when(
      max_proportion < 0.6 ~ "mixed",         
      sentiment[which.max(proportion)] == -1 ~ "risk", 
      sentiment[which.max(proportion)] == 1 ~ "opportunity"
    ),
    .groups = "drop"
  )

# Plot the US map with mixed sentiment
plot_usmap(data = state_sentiment, values = "sentiment_label", regions = "states") +
  scale_fill_manual(
    values = c(
      "opportunity" = "green", 
      "risk" = "red", 
      "mixed" = "purple"          
    ),
    name = "Sentiment",
    breaks = c("opportunity", "risk", "mixed"),
    labels = c("Opportunity", "Risk", "Mixed") 
  ) +
  labs(
    title = "Sentiment Distribution by State", 
    subtitle = "Opportunity, Risk, or Mixed Sentiment"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12)
  )

```

Wisconsin is split in terms of Senators and sentiment. South Dakota and Alaska also have split sentiment. 

```{r}

# Scatter plot for sentiment vs. ideology by party
ggplot(df, aes(x = ideology_score, y = sentiment, color = party)) +
  geom_jitter(alpha = 0.6, width = 0.1, height = 0.1) +  
  scale_color_manual(values = c("D" = "#00AEF3", "R" = "#E81B23", "I" = "#A45FE6")) +
  labs(
    title = "Sentiment vs. Ideology by Party",
    x = "Ideology Score (-1 = Liberal, 1 = Conservative)",
    y = "Sentiment (-1 = Risk, 1 = Opportunity)",
    color = "Party"
  ) +
  theme_minimal()

```

Democrats and Independents show a unified opportunity-oriented sentiment. Republicans exhibit a split, with a mix of opportunity and risk perceptions. 

```{r}

# Create correlation analysis between ideology_score and sentiment
ggplot(df, aes(x = ideology_score, y = sentiment)) +
  geom_point(aes(color = party), alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE) +
  scale_color_manual(values = c("D" = "#00AEF3", "R" = "#E81B23", "I" = "#A45FE6")) +
  labs(title = "Correlation between Ideology Score and Sentiment",
       x = "Ideology Score",
       y = "Sentiment",
       color = "Party") +
  theme_minimal()

```

The strong linear trend highlights the polarization between liberals and conservatives regarding climate sentiment. The presence of some conservative senators with opportunity-oriented sentiment (red points at Y = 1) indicates areas for potential collaboration on climate policy.

## Region


```{r}

df %>%
  mutate(region = case_when(
    state %in% c("maine", "vermont", "new hampshire", "massachusetts", 
                 "rhode island", "connecticut", "new york", "pennsylvania", 
                 "new jersey", "maryland", "delaware") ~ "Northeast",
    state %in% c("florida", "georgia", "alabama", "south carolina", 
                 "north carolina", "virginia", "west virginia", "kentucky",
                 "tennessee", "mississippi", "louisiana", "arkansas", "texas", "oklahoma") ~ "South",
    state %in% c("north dakota", "south dakota", "nebraska", "kansas", 
                 "minnesota", "iowa", "missouri", "wisconsin", "illinois", 
                 "indiana", "michigan", "ohio") ~ "Midwest",
    state %in% c("washington", "oregon", "california", "nevada", "idaho", 
                 "montana", "wyoming", "utah", "colorado", "arizona", 
                 "new mexico", "alaska", "hawaii") ~ "West"

```

```{r}

  )) %>%
  group_by(region, party) %>%
  summarise(count = n(), .groups = "drop") %>%
  ggplot(aes(x = region, y = count, fill = party)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_fill_manual(values = c("R" = "#E81B23", "D" = "#00AEF3", "I" = "#A45FE6")) +
  labs(
    title = "Party Composition by Region",
    subtitle = "Distribution of Political Parties Across US Regions",
    x = "Region",
    y = "Proportion",
    fill = "Party"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    legend.position = "bottom"
  )

```

```{r}

df %>%
  mutate(region = case_when(
    state %in% c("maine", "vermont", "new hampshire", "massachusetts", 
                 "rhode island", "connecticut", "maryland", "delaware", "new york", "pennsylvania", 
                 "new jersey") ~ "Northeast",
    state %in% c("florida", "georgia", "alabama", "south carolina", 
                 "north carolina", "virginia", "west virginia", "kentucky",
                 "tennessee", "mississippi", "louisiana", "arkansas", "oklahoma", "texas") ~ "South",
    state %in% c("north dakota", "south dakota", "nebraska", "kansas", 
                 "minnesota", "iowa", "missouri", "wisconsin", "illinois", 
                 "indiana", "michigan", "ohio") ~ "Midwest",
    state %in% c("washington", "oregon", "california", "nevada", "idaho", 
                 "montana", "wyoming", "utah", "colorado", "arizona", 
                 "new mexico", "alaska", "hawaii") ~ "West"
  )) %>%
  mutate(sentiment_label = case_when(
    sentiment == -1 ~ "Risk",
    sentiment == 1 ~ "Opportunity",
    TRUE ~ "Other"
  )) %>%
  group_by(region, sentiment_label) %>%
  summarise(count = n(), .groups = "drop") %>%
  ggplot(aes(x = region, y = count, fill = sentiment_label)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_fill_manual(values = c(
    "Risk" = "#E74C3C",
    "Opportunity" = "#2ECC71",
    "Other" = "#95A5A6"
  )) +
  labs(
    title = "Risk-Opportunity Distribution by Region",
    subtitle = "Distribution of Risk (-1) vs Opportunity (1) Sentiment",
    x = "Region",
    y = "Proportion",
    fill = "Assessment"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    legend.position = "bottom"
  )

```

## Statistical Tests

Null Hypothesis: There is no association between a senator's party affiliation (Democrat, Independent, Republican) and their sentiment (Risk, Opportunity).

Alternate Hypothesis: There is a significant association between a senator's party affiliation (Democrat, Independent, Republican) and their sentiment (Risk, Opportunity).

```{r}

# Create a contingency table for Party and Sentiment
party_sentiment_table <- table(df$party, df$sentiment)

# Perform a permutation test
perm_test_party <- chisq_test(party_sentiment_table, distribution = approximate(B = 10000))

# Print results
print("Permutation test for Party and Sentiment:")
print(perm_test_party)

```

chi-squared: 190.38
p-val: 0.0004
Since the p-value is < 0.05, we reject the null hypothesis and conclude that party affiliation is significantly associated with sentiment.

```{r}

# Create a contingency table for Region and Sentiment
region_sentiment_table <- table(df$region, df$sentiment)

# Perform a permutation test
region_perm_test <- chisq_test(region_sentiment_table, distribution = approximate(B = 10000))

# Print results
print("Permutation test for Region and Sentiment:")
print(region_perm_test)

```



```{r}

ggplot(df, aes(x = party, y = ideology_score, fill = party)) +
  geom_violin(alpha = 0.7) +
  geom_boxplot(width = 0.2, alpha = 0.7) +
  scale_fill_manual(values = c("D" = "#00AEF3", 
                              "R" = "#E81B23", 
                              "I" = "#A45FE6")) +
  labs(title = "Distribution of Ideology Scores by Party",
       x = "Party",
       y = "Ideology Score (-1 = Liberal, 1 = Conservative)") +
  theme_minimal()

```

