---
title: "Exploratory Data Analysis"
author: “Kristin Lloyd”
format: 
  html:
    toc: true
    embed-resources: true
    code-fold: true
---

```{r}

options(repos = c(CRAN = "https://cran.rstudio.com"))
install.packages("tidyverse")
install.packages("usmap")
install.packages("networkD3")
install.packages("coin")
library(tidyverse)
library(tidyr)
library(reshape2)
library(usmap)
library(networkD3)
library(coin)
library(dplyr)
library(ggplot2)

```

Let's load my processed dataset and look at my model labels. My main interpretations are in exploratory_data_analysis.html. The main point of this document is to show the differences between the model labeled visualizations and the indepdently labeled visualizations.

```{r}

df <- read.csv("../Data/processed-data/Senators_sentiment.csv")
head(df)

```

This dataset has a model_label column from my sentiment model. Let's see how the model's predictions effect my visualizations. 

```{r}

independent_counts <- df %>%
  filter(!is.na(model_label)) %>%
  count(model_label)

independent_counts <- independent_counts %>%
  mutate(model_label = factor(model_label, levels = c(-1, 1), labels = c("Risk", "Opportunity")))

ggplot(independent_counts, aes(x = model_label, y = n, fill = model_label)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("Risk" = "red", "Opportunity" = "#078d0f")) +
  labs(
    title = "Distribution of Sentiments",
    x = "Sentiment",
    y = "Count",
    fill = "Sentiment Type"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

As shown in my indepdent label visualization, opportunity sentiment is more prevalent than risk sentiment.

```{r}

grouped_data <- df %>%
  filter(!is.na(model_label)) %>%
  group_by(party, model_label) %>%
  summarise(Count = n(), .groups = 'drop') %>%
  group_by(party) %>%
  mutate(Proportion = Count / sum(Count)) 

grouped_data <- grouped_data %>%
  mutate(model_label = factor(model_label, levels = c(-1, 1), labels = c("Risk", "Opportunity")))

# Mosaic Plot
ggplot(grouped_data, aes(x = party, y = Proportion, fill = model_label)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_fill_manual(values = c("Risk" = "red", "Opportunity" = "#078d0f")) +
  labs(
    title = "Proportional Distribution of Sentiments by Party",
    x = "Party",
    y = "Proportion",
    fill = "Sentiment Type"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 0),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5)
  )

```

```{r}

flows_party <- df %>%
  filter(party %in% c("R", "I","D")) %>% 
  mutate(
    sentiment_label = ifelse(model_label == 1, "Opportunity", "Risk") 
  ) %>%
  group_by(party, sentiment_label) %>%
  summarise(value = n(), .groups = "drop") 

nodes_party <- data.frame(
  name = c(
    unique(flows_party$party),  
    unique(flows_party$sentiment_label)  
  )
)

links_party <- flows_party %>%
  mutate(
    source = match(party, nodes_party$name) - 1, 
    target = match(sentiment_label, nodes_party$name) - 1,  
    value = value  
  )

sankeyNetwork(
  Links = links_party, 
  Nodes = nodes_party,
  Source = "source",
  Target = "target",
  Value = "value", 
  NodeID = "name",
  sinksRight = TRUE, 
  nodeWidth = 30,
  fontSize = 12, 
  height = 600,
  width = 1000
)

```

These two plots mostly agree with the indepdent label plots. However, Democrats now have some risky sentiment and Republicans have a much larger split.

```{r}

bins <- c(-1.0, -0.8, -0.6, -0.4, -0.2, 0.0, 0.2, 0.4, 0.6, 0.8, 1.0)
labels <- c(
  "-1.0 to -0.8", "-0.8 to -0.6", "-0.6 to -0.4", "-0.4 to -0.2",
  "-0.2 to 0.0", "0.0 to 0.2", "0.2 to 0.4", "0.4 to 0.6",
  "0.6 to 0.8", "0.8 to 1.0"
)

grouped_data <- df %>%
  filter(!is.na(model_label)) %>%
  mutate(
    ideology_bin = cut(ideology_score, breaks = bins, labels = labels, include.lowest = TRUE),
    model_label = factor(model_label, levels = c(-1, 1), labels = c("Risk", "Opportunity"))
  ) %>%
  group_by(ideology_bin, model_label) %>%
  summarise(Count = n(), .groups = 'drop') %>%
  group_by(ideology_bin) %>%
  mutate(Proportion = Count / sum(Count))

# Mosaic Plot for Ideology Bins
ggplot(grouped_data, aes(x = ideology_bin, y = Proportion, fill = model_label)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_fill_manual(values = c("Risk" = "red", "Opportunity" = "#078d0f")) +
  labs(
    title = "Proportional Distribution of Sentiments by Ideology Bins",
    x = "Ideology Bin",
    y = "Proportion",
    fill = "Sentiment Type"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5)
  )

```

```{r}

heatmap_data <- df %>%
  mutate(
    ideology_bin = cut(ideology_score, 
                      breaks = seq(-1, 1, by = 0.2),
                      labels = c("-1.0 to -0.8", 
                               "-0.8 to -0.6", 
                               "-0.6 to -0.4", 
                               "-0.4 to -0.2", 
                               "-0.2 to 0.0",
                               "0.0 to 0.2",
                               "0.2 to 0.4",
                               "0.4 to 0.6",
                               "0.6 to 0.8",
                               "0.8 to 1.0")),
    model_label = ifelse(model_label == 1, "Opportunity", "Risk")
  ) %>%
  group_by(ideology_bin, model_label) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(ideology_bin) %>%
  mutate(percentage = (count / sum(count)) * 100) %>%
  ungroup()

heatmap_data <- heatmap_data %>%
  filter(!is.na(ideology_bin))

# Heatmap
ggplot(heatmap_data, aes(x = model_label, y = ideology_bin, fill = percentage)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.1f", percentage)), 
            color = "black", size = 4) +
  scale_fill_gradientn(
    colors = c("darkred", "red", "yellow", "lightgreen", "darkgreen"),  
    values = scales::rescale(c(0, 25, 50, 75, 100))  
  ) +
  labs(
    title = "Heatmap of Ideology Score Ranges by Sentiment (Risk and Opportunity)",
    x = "Sentiment",
    y = "Ideology Score Range",
    fill = "Percentage"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    axis.text.y = element_text(hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )

```

The heatmap does hold the relationship held in the indepdent label plot. However, in the indepdent plot, 0.2 to 0.4 was all green. Another big difference is the 0.8 to 1.0 showing a 33/66 split whereas the indepdent labels showed a 0/100 split. The model's visualizations show that the top conservative score is not associated with the most risky sentiment.

To test this relationship between ideology and sentiment, a correlation analysis would be beneficial. 

```{r}

# Correlation test between ideology score and sentiment
ideology_correlation <- cor.test(df$ideology_score, df$model_label)
print("\nCorrelation between Ideology Score and Sentiment:")
print(ideology_correlation)

```

The result of the correlation test is -0.5686831. This is less strong than the -0.8 value I got in my independent label analysis. This means that the model did not produce as telling of a model. 

Let's do some analysis by region now, starting with some maps. 

```{r}

df <- read.csv("../Data/processed-data/Senators_sentiment.csv")

state_sentiment <- df %>%
  filter(!is.na(model_label)) %>%               
  group_by(state, model_label) %>%
  summarise(count = n(), .groups = "drop") %>% 
  group_by(state) %>%
  mutate(
    total = sum(count),                        
    proportion = count / total              
  ) %>%
  summarise(
    max_proportion = max(proportion),       
    sentiment_label = case_when(
      max_proportion < 0.6 ~ "mixed",         
      model_label[which.max(proportion)] == -1 ~ "risk", 
      model_label[which.max(proportion)] == 1 ~ "opportunity"
    ),
    .groups = "drop"
  )

# Plot the US map with mixed sentiment
plot_usmap(data = state_sentiment, values = "sentiment_label", regions = "states") +
  scale_fill_manual(
    values = c(
      "opportunity" = "green", 
      "risk" = "red", 
      "mixed" = "purple"          
    ),
    name = "Sentiment",
    breaks = c("opportunity", "risk", "mixed"),
    labels = c("Opportunity", "Risk", "Mixed") 
  ) +
  labs(
    title = "Sentiment Distribution by State", 
    subtitle = "Opportunity, Risk, or Mixed Sentiment"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12)
  )

```

Texas, Wisconsin, Iowa, Tennessee, Oklahoma, South Dakota, and Colorado now have mixed sentiment. Wisconsin and South Dakota are the only state that previously had mixed sentiment out of these states. Based on independent labels, Pennsylvania, Alaska, and South Dakota should have mixed sentiment. The model's labels also classified a lot of states as having opportunistic sentiment when the indepdent labels claim that it should be negative sentiment. 

```{r}

# Region and Party
party_region_counts <- df %>%
  group_by(region, party) %>%
  summarise(Count = n(), .groups = "drop")

ggplot(party_region_counts, aes(x = region, y = Count, fill = party)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("D" = "blue", "R" = "red", "I" = "purple")) +
  labs(
    title = "Party Composition by Region",
    x = "Region",
    y = "Count",
    fill = "Party"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```

This relationship is still preserved compared to the indepdently labeled model. 

```{r}

# Region and Sentiment
sentiment_region_props <- df %>%
  group_by(region, model_label) %>%
  summarise(Count = n(), .groups = "drop") %>%
  group_by(region) %>%
  mutate(Proportion = Count / sum(Count))

ggplot(sentiment_region_props, aes(x = region, y = Proportion, fill = as.factor(model_label))) +
  geom_bar(stat = "identity", position = "fill") +
  scale_fill_manual(values = c("-1" = "red", "1" = "#078d0f"), labels = c("Risk", "Opportunity")) +
  labs(
    title = "Sentiment Distribution by Region",
    x = "Region",
    y = "Proportion",
    fill = "Sentiment"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```

The relationships are still preserved.

```{r}

heatmap_data <- df %>%
  mutate(
    model_label = ifelse(model_label == 1, "Opportunity", "Risk")  
  ) %>%
  group_by(region, model_label) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(region) %>%
  mutate(percentage = (count / sum(count)) * 100) %>%
  ungroup()

# Heatmap
ggplot(heatmap_data, aes(x = model_label, y = region, fill = percentage)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.1f%%", percentage)), 
            color = "black", size = 4) +
  scale_fill_gradientn(
    colors = c("darkred", "red", "yellow", "lightgreen", "darkgreen"),  
    values = scales::rescale(c(0, 25, 50, 75, 100)), 
    name = "Percentage"
  ) +
  labs(
    title = "Heatmap of Sentiment Distribution by Region",
    x = "Sentiment",
    y = "Region"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    axis.text.y = element_text(hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )

```

```{r}

flows_region <- df %>%
  mutate(
    sentiment_label = ifelse(model_label == 1, "Opportunity", "Risk")  
  ) %>%
  group_by(region, sentiment_label) %>%
  summarise(value = n(), .groups = "drop") 

nodes_region <- data.frame(
  name = c(
    unique(flows_region$region),     
    unique(flows_region$sentiment_label)  
  )
)

links_region <- flows_region %>%
  mutate(
    source = match(region, nodes_region$name) - 1,       
    target = match(sentiment_label, nodes_region$name) - 1,  
    value = value                                   
  )

sankeyNetwork(
  Links = links_region, 
  Nodes = nodes_region,
  Source = "source",
  Target = "target",
  Value = "value", 
  NodeID = "name",
  sinksRight = TRUE, 
  nodeWidth = 30,
  fontSize = 12, 
  height = 600,
  width = 1000
)

```

This heatmap is not the same compared to the indepdent label plot. The only region that preserves the relationship is Northeast. 

## Conclusion

The analysis reveals a divide in how U.S. senators view climate change. Democrats and Independents tend to see it in an opportunistic way, while Republicans are more likely to view it in a risky way. This divide is reflected in the negative correlation between a senator's ideology and their climate sentiment. The regional analysis shows the Northeast and West are more opportunity-focused, while the Midwest and South have a mix of risky and opportunistic perspectives. Some states have a more balanced, "mixed" sentiment, suggesting potential for bipartisan cooperation.
