---
title: "Exploratory Data Analysis"
author: “Kristin Lloyd”
format: 
  html:
    toc: true
    embed-resources: true
    code-fold: true
---

```{r}

options(repos = c(CRAN = "https://cran.rstudio.com"))
install.packages("tidyverse")
install.packages("usmap")
install.packages("networkD3")
install.packages("coin")
library(tidyverse)
library(tidyr)
library(reshape2)
library(usmap)
library(networkD3)
library(coin)
library(dplyr)
library(ggplot2)

```

```{r}

df <- read.csv("Data/processed-data/Senators_sentiment.csv")

```

```{r}

senator_summary <- df %>%
  mutate(full_name = paste(first, last)) %>%  
  distinct(state, full_name, party) %>%  
  arrange(state, full_name) %>%             
  group_by(state) %>%
  summarise(
    Senators = paste(paste(full_name, paste0("(", party, ")"), sep = " "), 
                     collapse = ", "),       
    Count = n()               
  ) %>%
  ungroup()
  
print(senator_summary, n = Inf)

```

Each state except for Ohio has two senators. Ohio currently has one senator because J.D. Vance stepped down to become the vice president. So we are analyzing 99 senators as of November 30th, 2024.

```{r}

independent_counts <- df %>%
  filter(!is.na(sentiment)) %>%
  count(sentiment)

independent_counts <- independent_counts %>%
  mutate(sentiment = factor(sentiment, levels = c(-1, 1), labels = c("Risk", "Opportunity")))

ggplot(independent_counts, aes(x = sentiment, y = n, fill = sentiment)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("Risk" = "red", "Opportunity" = "#078d0f")) +
  labs(
    title = "Distribution of Sentiments",
    x = "Sentiment",
    y = "Count",
    fill = "Sentiment Type"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

Let's first look at how much risk and opportunity sentiment are in our dataset regarding climate change. Opportunity sentiment is more prevalent than risk sentiment, indicating that the senate generally has a proactive stance on addressing challenges and pursuing solutions. Now let's look at a proportional distribution of sentiment by party. 

```{r}

grouped_data <- df %>%
  filter(!is.na(sentiment)) %>%
  group_by(party, sentiment) %>%
  summarise(Count = n(), .groups = 'drop') %>%
  group_by(party) %>%
  mutate(Proportion = Count / sum(Count)) 

grouped_data <- grouped_data %>%
  mutate(sentiment = factor(sentiment, levels = c(-1, 1), labels = c("Risk", "Opportunity")))

# Mosaic Plot
ggplot(grouped_data, aes(x = party, y = Proportion, fill = sentiment)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_fill_manual(values = c("Risk" = "red", "Opportunity" = "#078d0f")) +
  labs(
    title = "Proportional Distribution of Sentiments by Party",
    x = "Party",
    y = "Proportion",
    fill = "Sentiment Type"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 0),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5)
  )

```

```{r}

flows_party <- df %>%
  filter(party %in% c("R", "D")) %>% 
  mutate(
    sentiment_label = ifelse(sentiment == 1, "Opportunity", "Risk") 
  ) %>%
  group_by(party, sentiment_label) %>%
  summarise(value = n(), .groups = "drop") 

nodes_party <- data.frame(
  name = c(
    unique(flows_party$party),  
    unique(flows_party$sentiment_label)  
  )
)

links_party <- flows_party %>%
  mutate(
    source = match(party, nodes_party$name) - 1, 
    target = match(sentiment_label, nodes_party$name) - 1,  
    value = value  
  )

sankeyNetwork(
  Links = links_party, 
  Nodes = nodes_party,
  Source = "source",
  Target = "target",
  Value = "value", 
  NodeID = "name",
  sinksRight = TRUE, 
  nodeWidth = 30,
  fontSize = 12, 
  height = 600,
  width = 1000
)

```

Democratic and Indepdent senators overwhelmingly view climate change as a very real challenge that they want to tackle. Republicans introduce risk while talking about climate change, meaning they do not believe in it or do not support tackling it. However, there are different levels of being a Democrat and being a Republican. Our senators have been given ideology rankings. Let's analyze how sentiment changes as ideology ranks do. The ranks go from -1 to 1, with -1 being very librtal and 1 being very conservative. 

```{r}

ggplot(df, aes(x = party, y = ideology_score, fill = party)) +
  geom_violin(alpha = 0.7) +
  geom_boxplot(width = 0.2, alpha = 0.7) +
  scale_fill_manual(values = c("D" = "#00AEF3", 
                              "R" = "#E81B23", 
                              "I" = "#A45FE6")) +
  labs(title = "Distribution of Ideology Scores by Party",
       x = "Party",
       y = "Ideology Score (-1 = Liberal, 1 = Conservative)") +
  theme_minimal()

```

Interpretation:

```{r}

# Create ideology bins
bins <- c(-1.0, -0.8, -0.6, -0.4, -0.2, 0.0, 0.2, 0.4, 0.6, 0.8, 1.0)
labels <- c(
  "-1.0 to -0.8", "-0.8 to -0.6", "-0.6 to -0.4", "-0.4 to -0.2",
  "-0.2 to 0.0", "0.0 to 0.2", "0.2 to 0.4", "0.4 to 0.6",
  "0.6 to 0.8", "0.8 to 1.0"
)

# Add ideology bins and calculate proportions
grouped_data <- df %>%
  filter(!is.na(sentiment)) %>%
  mutate(
    ideology_bin = cut(ideology_score, breaks = bins, labels = labels, include.lowest = TRUE),
    sentiment = factor(sentiment, levels = c(-1, 1), labels = c("Risk", "Opportunity"))
  ) %>%
  group_by(ideology_bin, sentiment) %>%
  summarise(Count = n(), .groups = 'drop') %>%
  group_by(ideology_bin) %>%
  mutate(Proportion = Count / sum(Count))

# Mosaic Plot for Ideology Bins
ggplot(grouped_data, aes(x = ideology_bin, y = Proportion, fill = sentiment)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_fill_manual(values = c("Risk" = "red", "Opportunity" = "#078d0f")) +
  labs(
    title = "Proportional Distribution of Sentiments by Ideology Bins",
    x = "Ideology Bin",
    y = "Proportion",
    fill = "Sentiment Type"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5)
  )

```

```{r}

heatmap_data <- df %>%
  mutate(
    ideology_bin = cut(ideology_score, 
                      breaks = seq(-1, 1, by = 0.2),
                      labels = c("-1.0 to -0.8", 
                               "-0.8 to -0.6", 
                               "-0.6 to -0.4", 
                               "-0.4 to -0.2", 
                               "-0.2 to 0.0",
                               "0.0 to 0.2",
                               "0.2 to 0.4",
                               "0.4 to 0.6",
                               "0.6 to 0.8",
                               "0.8 to 1.0")),
    sentiment = ifelse(sentiment == 1, "Opportunity", "Risk")
  ) %>%
  group_by(ideology_bin, sentiment) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(ideology_bin) %>%
  mutate(percentage = (count / sum(count)) * 100) %>%
  ungroup()

heatmap_data <- heatmap_data %>%
  filter(!is.na(ideology_bin))

# Heatmap
ggplot(heatmap_data, aes(x = sentiment, y = ideology_bin, fill = percentage)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.1f", percentage)), 
            color = "black", size = 4) +
  scale_fill_gradientn(
    colors = c("darkred", "red", "yellow", "lightgreen", "darkgreen"),  
    values = scales::rescale(c(0, 25, 50, 75, 100))  
  ) +
  labs(
    title = "Heatmap of Ideology Score Ranges by Sentiment (Risk and Opportunity)",
    x = "Sentiment",
    y = "Ideology Score Range",
    fill = "Percentage"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    axis.text.y = element_text(hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )

```

-1 to 0 ideology score were all opportunity sentiment. Negative sentiment began dominating more and more as we approach positive 1. Let's do a correlation test to confirm this result.

```{r}

# Correlation test between ideology score and sentiment
ideology_correlation <- cor.test(df$ideology_score, df$sentiment)
print("\nCorrelation between Ideology Score and Sentiment:")
print(ideology_correlation)

```

The correlation value is -0.8134339! However, let's also look at this scatter plot.

```{r}

# Scatter plot for sentiment vs. ideology by party
ggplot(df, aes(x = ideology_score, y = sentiment, color = party)) +
  geom_jitter(alpha = 0.6, width = 0.1, height = 0.1) +  
  scale_color_manual(values = c("D" = "#00AEF3", "R" = "#E81B23", "I" = "#A45FE6")) +
  labs(
    title = "Sentiment vs. Ideology by Party",
    x = "Ideology Score (-1 = Liberal, 1 = Conservative)",
    y = "Sentiment (-1 = Risk, 1 = Opportunity)",
    color = "Party"
  ) +
  theme_minimal()

```

We can see here that some Republican's do have positive sentiment regarding climate change. Democrats and Independents show a unified opportunity-oriented sentiment. Republicans exhibit a split, with a mix of opportunity and risk perceptions. 

Let's do some analysis by region now, starting with some maps. 

```{r}

df <- df %>%
  mutate(state = tolower(state.name[match(state, state.abb)])) 

state_senators <- df %>%
  distinct(state, party) %>%                           
  group_by(state) %>%
  summarise(
    party_composition = case_when(
      all(party == "R") ~ "Republican",
      all(party == "D") ~ "Democrat",
      any(party == "I") & all(party %in% c("I", "D")) ~ "Split (Independent & Democrat)", 
      any(party == "I") & all(party %in% c("I", "R")) ~ "Split (Independent & Republican)",
      TRUE ~ "Split (Republican & Democrat)"
    ),
    .groups = "drop"
  )

party_colors <- c(
  "Republican" = "red",
  "Democrat" = "blue",
  "Split (Independent & Democrat)" = "yellow",
  "Split (Independent & Republican)" = "orange",
  "Split (Republican & Democrat)" = "purple"
)

plot_usmap(data = state_senators, values = "party_composition", regions = "states") +
  scale_fill_manual(
    values = party_colors,
    name = "Party Composition"
  ) +
  labs(
    title = "Senator Party Composition by State",
    subtitle = "Republican, Democrat, Independent, or Split"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    panel.background = element_rect(fill = "white"),
    panel.grid = element_blank()
  )

```

Most states elect senators from the same party, with the exception of Wisconsin, Pennsylvania, Maine, and Vermont. 

```{r}

df <- df %>%
  mutate(state = tolower(state.name[match(state, state.abb)])) 

state_senators <- df %>%
  group_by(state) %>%
  summarise(
    avg_ideology_score = mean(ideology_score, na.rm = TRUE),  
    .groups = "drop"
  )

plot_usmap(data = state_senators, values = "avg_ideology_score", regions = "states") +
  scale_fill_gradient2(
    low = "#3097ff",   # Liberal (-1)
    mid = "white",  # Neutral (0)
    high = "red",    # Conservative (1)
    midpoint = 0,    # Center the gradient around 0
    name = "Avg. Ideology Score"
  ) +
  labs(
    title = "Average Ideology Score of Senators by State",
    subtitle = "Blue = More Liberal, Red = More Conservative"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    panel.background = element_rect(fill = "white"),
    panel.grid = element_blank()
  )

```

The map shows the average ideology score of senators by state, with blue indicating liberal states, red indicating conservative states, and white representing neutral ideology. A majority of states seem to be moderate deocrat or moderate republican. There are only a few states with brighter blue/red values. 

```{r}

state_sentiment <- df %>%
  filter(!is.na(sentiment)) %>%               
  group_by(state, sentiment) %>%
  summarise(count = n(), .groups = "drop") %>% 
  group_by(state) %>%
  mutate(
    total = sum(count),                        
    proportion = count / total              
  ) %>%
  summarise(
    max_proportion = max(proportion),       
    sentiment_label = case_when(
      max_proportion < 0.6 ~ "mixed",         
      sentiment[which.max(proportion)] == -1 ~ "risk", 
      sentiment[which.max(proportion)] == 1 ~ "opportunity"
    ),
    .groups = "drop"
  )

# Plot the US map with mixed sentiment
plot_usmap(data = state_sentiment, values = "sentiment_label", regions = "states") +
  scale_fill_manual(
    values = c(
      "opportunity" = "green", 
      "risk" = "red", 
      "mixed" = "purple"          
    ),
    name = "Sentiment",
    breaks = c("opportunity", "risk", "mixed"),
    labels = c("Opportunity", "Risk", "Mixed") 
  ) +
  labs(
    title = "Sentiment Distribution by State", 
    subtitle = "Opportunity, Risk, or Mixed Sentiment"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12)
  )

```

Wisconsin had mixed senators and had an average ideology score of around 0 since it was shaded white in the previous plot. On this plot, we see that it has mixed sentiment which makes sense. Alaska has mixed sentiment because despite both senators being republican, 
Iowa has mixed sentiment because 
North Carolina has mixed sentiment because

Let's run a chi square test to determine if there is a significant relationship between ideology score and sentiment. 

Null Hypothesis: There is no association between a senator's party affiliation (Democrat, Independent, Republican) and their sentiment (Risk, Opportunity).

Alternate Hypothesis: There is a significant association between a senator's party affiliation (Democrat, Independent, Republican) and their sentiment (Risk, Opportunity).

```{r}

install.packages("coin")  
library(coin) 

party_sentiment_table <- table(df$party, df$sentiment)

perm_test_party <- chisq_test(party_sentiment_table, distribution = approximate(B = 10000))

print("Permutation test for Party and Sentiment:")
print(perm_test_party)

```

chi-squared: 190.38
p-val: 0.0004
Since the p-value is < 0.05, we reject the null hypothesis and conclude that party affiliation is significantly associated with sentiment.

Let's separate into regions now. 

```{r}

# Region and Party
party_region_counts <- df %>%
  group_by(region, party) %>%
  summarise(Count = n(), .groups = "drop")

ggplot(party_region_counts, aes(x = region, y = Count, fill = party)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("D" = "blue", "R" = "red", "I" = "purple")) +
  labs(
    title = "Party Composition by Region",
    x = "Region",
    y = "Count",
    fill = "Party"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```

Northeast and West regions are more Democratic than Republican and Midwest and South regions are more Republican than Democratic.

```{r}

# Region and Sentiment
sentiment_region_props <- df %>%
  group_by(region, sentiment) %>%
  summarise(Count = n(), .groups = "drop") %>%
  group_by(region) %>%
  mutate(Proportion = Count / sum(Count))

ggplot(sentiment_region_props, aes(x = region, y = Proportion, fill = as.factor(sentiment))) +
  geom_bar(stat = "identity", position = "fill") +
  scale_fill_manual(values = c("-1" = "red", "1" = "#078d0f"), labels = c("Risk", "Opportunity")) +
  labs(
    title = "Sentiment Distribution by Region",
    x = "Region",
    y = "Proportion",
    fill = "Sentiment"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```

Northeast and West regions are more opportunity than risk and Midwest and South regions are more risk than opportunity.

```{r}

# Boxplot for ideology score by region
ggplot(df, aes(x = region, y = ideology_score, fill = region)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_brewer(palette = "Set3") +
  labs(
    title = "Ideology Score Distribution by Region",
    x = "Region",
    y = "Ideology Score (-1 = Liberal, 1 = Conservative)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )

```

The Northeast is predominantly liberal with minimal variation, while the West and Midwest exhibit the greatest ideological diversity, spanning from strongly liberal to strongly conservative. The South leans slightly conservative but is less polarized compared to other regions.

```{r}

heatmap_data <- df %>%
  mutate(
    sentiment = ifelse(sentiment == 1, "Opportunity", "Risk")  
  ) %>%
  group_by(region, sentiment) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(region) %>%
  mutate(percentage = (count / sum(count)) * 100) %>%
  ungroup()

# Heatmap
ggplot(heatmap_data, aes(x = sentiment, y = region, fill = percentage)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.1f%%", percentage)), 
            color = "black", size = 4) +
  scale_fill_gradientn(
    colors = c("darkred", "red", "yellow", "lightgreen", "darkgreen"),  
    values = scales::rescale(c(0, 25, 50, 75, 100)), 
    name = "Percentage"
  ) +
  labs(
    title = "Heatmap of Sentiment Distribution by Region",
    x = "Sentiment",
    y = "Region"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    axis.text.y = element_text(hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )

```

The Northeast is the most optimistic region, strongly favoring opportunity. The Midwest is the most divided, highlighting significant differences in sentiment within the region. The West and South share similar sentiment distributions, leaning optimistic but still spreading risk regarding climate change.

```{r}

flows_region <- df %>%
  mutate(
    sentiment_label = ifelse(sentiment == 1, "Opportunity", "Risk")  
  ) %>%
  group_by(region, sentiment_label) %>%
  summarise(value = n(), .groups = "drop") 

nodes_region <- data.frame(
  name = c(
    unique(flows_region$region),     
    unique(flows_region$sentiment_label)  
  )
)

links_region <- flows_region %>%
  mutate(
    source = match(region, nodes_region$name) - 1,       
    target = match(sentiment_label, nodes_region$name) - 1,  
    value = value                                   
  )

sankeyNetwork(
  Links = links_region, 
  Nodes = nodes_region,
  Source = "source",
  Target = "target",
  Value = "value", 
  NodeID = "name",
  sinksRight = TRUE, 
  nodeWidth = 30,
  fontSize = 12, 
  height = 600,
  width = 1000
)

```

Let's run a correlation test.

```{r}

# Correlation test between ideology score and sentiment
ideology_correlation <- cor.test(df$ideology_score, df$sentiment)
print("\nCorrelation between Ideology Score and Sentiment:")
print(ideology_correlation)

```

The value was -0.8134339

Let's run a chi square test to determine if there is a significant relationship between region and sentiment. 

Null hypothesis: There is no association between a senator's region (e.g., Midwest, South, West, Northeast) and their sentiment (Risk, Opportunity).

Alternate hypothesis: There is a significant association between a senator's region and their sentiment.

```{r}

region_sentiment_table <- table(df$region, df$sentiment)

perm_test_region <- chisq_test(region_sentiment_table, distribution = approximate(B = 10000))

print("Permutation test for Region and Sentiment:")
print(perm_test_region)

```

chi square: 69.002
p-val: 0.0004
We reject the null hypothesis and conclude that region is significantly associated with sentiment. 



```{r}

flows_region <- df %>%
  mutate(
    sentiment_label = ifelse(sentiment == 1, "Opportunity", "Risk")  
  ) %>%
  group_by(region, sentiment_label) %>%
  summarise(value = n(), .groups = "drop") 

nodes_region <- data.frame(
  name = c(
    unique(flows_region$region),     
    unique(flows_region$sentiment_label)  
  )
)

links_region <- flows_region %>%
  mutate(
    source = match(region, nodes_region$name) - 1,       
    target = match(sentiment_label, nodes_region$name) - 1,  
    value = value                                   
  )

sankeyNetwork(
  Links = links_region, 
  Nodes = nodes_region,
  Source = "source",
  Target = "target",
  Value = "value", 
  NodeID = "name",
  sinksRight = TRUE, 
  nodeWidth = 30,
  fontSize = 12, 
  height = 600,
  width = 1000
)

```