---
title: "Exploratory Data Analysis"
author: “Kristin Lloyd”
format: 
  html:
    toc: true
    embed-resources: true
    code-fold: true
---

```{r}

options(repos = c(CRAN = "https://cran.rstudio.com"))
install.packages("tidyverse")
install.packages("usmap")
install.packages("networkD3")
install.packages("coin")
library(tidyverse)
library(tidyr)
library(reshape2)
library(usmap)
library(networkD3)
library(coin)
library(dplyr)
library(ggplot2)

```

Let's load my processed dataset and look at my model labels. 

```{r}

df <- read.csv("Data/processed-data/Senators_sentiment.csv")
head(df)

```

Each state except for Ohio has two senators. Ohio currently has one senator because J.D. Vance stepped down to become the vice president. So we are analyzing 99 senators as of November 30th, 2024.

```{r}

independent_counts <- df %>%
  filter(!is.na(model_label)) %>%
  count(model_label)

independent_counts <- independent_counts %>%
  mutate(model_label = factor(model_label, levels = c(-1, 1), labels = c("Risk", "Opportunity")))

ggplot(independent_counts, aes(x = model_label, y = n, fill = model_label)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("Risk" = "red", "Opportunity" = "#078d0f")) +
  labs(
    title = "Distribution of Sentiments",
    x = "Sentiment",
    y = "Count",
    fill = "Sentiment Type"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

Let's first look at how much risk and opportunity sentiment are in our dataset regarding climate change. Opportunity sentiment is more prevalent than risk sentiment, indicating that the senate generally has a proactive stance on addressing challenges and pursuing solutions. Now let's look at a proportional distribution of sentiment by party. 

```{r}

grouped_data <- df %>%
  filter(!is.na(model_label)) %>%
  group_by(party, model_label) %>%
  summarise(Count = n(), .groups = 'drop') %>%
  group_by(party) %>%
  mutate(Proportion = Count / sum(Count)) 

grouped_data <- grouped_data %>%
  mutate(model_label = factor(model_label, levels = c(-1, 1), labels = c("Risk", "Opportunity")))

# Mosaic Plot
ggplot(grouped_data, aes(x = party, y = Proportion, fill = model_label)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_fill_manual(values = c("Risk" = "red", "Opportunity" = "#078d0f")) +
  labs(
    title = "Proportional Distribution of Sentiments by Party",
    x = "Party",
    y = "Proportion",
    fill = "Sentiment Type"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 0),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5)
  )

```

```{r}

flows_party <- df %>%
  filter(party %in% c("R", "I","D")) %>% 
  mutate(
    sentiment_label = ifelse(model_label == 1, "Opportunity", "Risk") 
  ) %>%
  group_by(party, sentiment_label) %>%
  summarise(value = n(), .groups = "drop") 

nodes_party <- data.frame(
  name = c(
    unique(flows_party$party),  
    unique(flows_party$sentiment_label)  
  )
)

links_party <- flows_party %>%
  mutate(
    source = match(party, nodes_party$name) - 1, 
    target = match(sentiment_label, nodes_party$name) - 1,  
    value = value  
  )

sankeyNetwork(
  Links = links_party, 
  Nodes = nodes_party,
  Source = "source",
  Target = "target",
  Value = "value", 
  NodeID = "name",
  sinksRight = TRUE, 
  nodeWidth = 30,
  fontSize = 12, 
  height = 600,
  width = 1000
)

```

Based on these two plots, democratic and indepdent senators overwhelmingly view climate change as a very real challenge that they want to tackle. Republicans generally introduce risk while talking about climate change, meaning they do not believe in it or do not support tackling it. However, there are different levels of being a Democrat and being a Republican. Our senators have been given ideology rankings. Let's analyze how sentiment changes as ideology ranks do. The ranks go from -1 to 1, with -1 being very liberal and 1 being very conservative. 

```{r}

ggplot(df, aes(x = party, y = ideology_score, fill = party)) +
  geom_violin(alpha = 0.7) +
  geom_boxplot(width = 0.2, alpha = 0.7) +
  scale_fill_manual(values = c("D" = "#00AEF3", 
                              "R" = "#E81B23", 
                              "I" = "#A45FE6")) +
  labs(title = "Distribution of Ideology Scores by Party",
       x = "Party",
       y = "Ideology Score (-1 = Liberal, 1 = Conservative)") +
  theme_minimal()

```

Democrat senators exhibit a more concentrated liberal ideology and Republicans displaying a wider range of ideology scores.

```{r}

medians <- df %>%
  group_by(party) %>%
  summarise(median_ideology_score = median(ideology_score, na.rm = TRUE))

print(medians)

```

The median ideology score for Democratic (D) senators is -0.357, which falls within the moderately liberal range. The median ideology score for Independent (I) senators is -0.338, which is very close to the Democratic senators and also falls within the moderately liberal range. The median ideology score for Republican (R) senators is 0.538, which is in the moderately conservative range.

```{r}

# Create ideology bins
bins <- c(-1.0, -0.8, -0.6, -0.4, -0.2, 0.0, 0.2, 0.4, 0.6, 0.8, 1.0)
labels <- c(
  "-1.0 to -0.8", "-0.8 to -0.6", "-0.6 to -0.4", "-0.4 to -0.2",
  "-0.2 to 0.0", "0.0 to 0.2", "0.2 to 0.4", "0.4 to 0.6",
  "0.6 to 0.8", "0.8 to 1.0"
)

# Add ideology bins and calculate proportions
grouped_data <- df %>%
  filter(!is.na(model_label)) %>%
  mutate(
    ideology_bin = cut(ideology_score, breaks = bins, labels = labels, include.lowest = TRUE),
    model_label = factor(model_label, levels = c(-1, 1), labels = c("Risk", "Opportunity"))
  ) %>%
  group_by(ideology_bin, model_label) %>%
  summarise(Count = n(), .groups = 'drop') %>%
  group_by(ideology_bin) %>%
  mutate(Proportion = Count / sum(Count))

# Mosaic Plot for Ideology Bins
ggplot(grouped_data, aes(x = ideology_bin, y = Proportion, fill = model_label)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_fill_manual(values = c("Risk" = "red", "Opportunity" = "#078d0f")) +
  labs(
    title = "Proportional Distribution of Sentiments by Ideology Bins",
    x = "Ideology Bin",
    y = "Proportion",
    fill = "Sentiment Type"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5)
  )

```

```{r}

heatmap_data <- df %>%
  mutate(
    ideology_bin = cut(ideology_score, 
                      breaks = seq(-1, 1, by = 0.2),
                      labels = c("-1.0 to -0.8", 
                               "-0.8 to -0.6", 
                               "-0.6 to -0.4", 
                               "-0.4 to -0.2", 
                               "-0.2 to 0.0",
                               "0.0 to 0.2",
                               "0.2 to 0.4",
                               "0.4 to 0.6",
                               "0.6 to 0.8",
                               "0.8 to 1.0")),
    model_label = ifelse(model_label == 1, "Opportunity", "Risk")
  ) %>%
  group_by(ideology_bin, model_label) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(ideology_bin) %>%
  mutate(percentage = (count / sum(count)) * 100) %>%
  ungroup()

heatmap_data <- heatmap_data %>%
  filter(!is.na(ideology_bin))

# Heatmap
ggplot(heatmap_data, aes(x = model_label, y = ideology_bin, fill = percentage)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.1f", percentage)), 
            color = "black", size = 4) +
  scale_fill_gradientn(
    colors = c("darkred", "red", "yellow", "lightgreen", "darkgreen"),  
    values = scales::rescale(c(0, 25, 50, 75, 100))  
  ) +
  labs(
    title = "Heatmap of Ideology Score Ranges by Sentiment (Risk and Opportunity)",
    x = "Sentiment",
    y = "Ideology Score Range",
    fill = "Percentage"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    axis.text.y = element_text(hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )

```

Senators with more liberal ideology scores overwhelmingly demonstrate an opportunity based sentiment. In contrast, as the ideology scores become more conservative, the risk sentiment starts to dominate, suggesting a more cautious or skeptical perspective. To confirm this relationship between ideology and sentiment, a correlation analysis would be beneficial. 

```{r}

# Correlation test between ideology score and sentiment
ideology_correlation <- cor.test(df$ideology_score, df$model_label)
print("\nCorrelation between Ideology Score and Sentiment:")
print(ideology_correlation)

```

The results of the correlation analysis provide strong evidence to support the observed relationship between ideology and sentiment. The correlation coefficient is -0.813, indicating a negative correlation between these two variables. 

Let's do some analysis by region now, starting with some maps. 

```{r}

df <- read.csv("../Data/processed-data/Senators_sentiment.csv")

library(dplyr)
library(usmap)
library(ggplot2)

df <- df %>%
  mutate(state = tolower(state.name[match(state, state.abb)]))

state_senators <- df %>%
  distinct(state, party) %>%
  group_by(state) %>%
  summarise(
    party_composition = case_when(
      all(party == "R") ~ "Republican",
      all(party == "D") ~ "Democrat",
      any(party == "I") & all(party %in% c("I", "D")) ~ "Split (Independent & Democrat)", 
      any(party == "I") & all(party %in% c("I", "R")) ~ "Split (Independent & Republican)",
      TRUE ~ "Split (Republican & Democrat)"
    ),
    .groups = "drop"
  )

state_senators <- state_senators[!is.na(state_senators$party_composition), ]

party_colors <- c(
  "Republican" = "red",
  "Democrat" = "blue",
  "Split (Independent & Democrat)" = "yellow",
  "Split (Independent & Republican)" = "orange",
  "Split (Republican & Democrat)" = "purple"
)

# Plot Map
plot_usmap(data = state_senators, values = "party_composition", regions = "states") +
  scale_fill_manual(
    values = party_colors,
    name = "Party Composition",
    na.value = "white"
  ) +
  labs(
    title = "Senator Party Composition by State",
    subtitle = "Republican, Democrat, Independent, or Split"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    panel.background = element_rect(fill = "white"),
    panel.grid = element_blank()
  )

```

Most states elect senators from the same party, with the exception of Wisconsin, Pennsylvania, Maine, and Vermont. 

```{r}

df <- read.csv("../Data/processed-data/Senators_sentiment.csv")

df <- df %>%
  mutate(state = tolower(state.name[match(state, state.abb)])) 

state_senators <- df %>%
  group_by(state) %>%
  summarise(
    avg_ideology_score = mean(ideology_score, na.rm = TRUE),  
    .groups = "drop"
  )

plot_usmap(data = state_senators, values = "avg_ideology_score", regions = "states") +
  scale_fill_gradient2(
    low = "#3097ff",   # Liberal (-1)
    mid = "white",  # Neutral (0)
    high = "red",    # Conservative (1)
    midpoint = 0,    # Center the gradient around 0
    name = "Avg. Ideology Score"
  ) +
  labs(
    title = "Average Ideology Score of Senators by State",
    subtitle = "Blue = More Liberal, Red = More Conservative"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    panel.background = element_rect(fill = "white"),
    panel.grid = element_blank()
  )

```

The map shows that most states have senators with more moderate ideological leanings, reflecting a balance of Democratic and Republican views. However, a few states have senators with more extreme liberal or conservative positions. This ideological polarization in those states could make it harder for their senators to find common ground and work together on climate change.  

```{r}

df <- read.csv("../Data/processed-data/Senators_sentiment.csv")

state_sentiment <- df %>%
  filter(!is.na(model_label)) %>%               
  group_by(state, model_label) %>%
  summarise(count = n(), .groups = "drop") %>% 
  group_by(state) %>%
  mutate(
    total = sum(count),                        
    proportion = count / total              
  ) %>%
  summarise(
    max_proportion = max(proportion),       
    sentiment_label = case_when(
      max_proportion < 0.6 ~ "mixed",         
      model_label[which.max(proportion)] == -1 ~ "risk", 
      model_label[which.max(proportion)] == 1 ~ "opportunity"
    ),
    .groups = "drop"
  )

# Plot the US map with mixed sentiment
plot_usmap(data = state_sentiment, values = "sentiment_label", regions = "states") +
  scale_fill_manual(
    values = c(
      "opportunity" = "green", 
      "risk" = "red", 
      "mixed" = "purple"          
    ),
    name = "Sentiment",
    breaks = c("opportunity", "risk", "mixed"),
    labels = c("Opportunity", "Risk", "Mixed") 
  ) +
  labs(
    title = "Sentiment Distribution by State", 
    subtitle = "Opportunity, Risk, or Mixed Sentiment"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12)
  )

```

You can see that it now classifies states wrong. This map is completely different from my first map.

```{r}

# Region and Party
party_region_counts <- df %>%
  group_by(region, party) %>%
  summarise(Count = n(), .groups = "drop")

ggplot(party_region_counts, aes(x = region, y = Count, fill = party)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("D" = "blue", "R" = "red", "I" = "purple")) +
  labs(
    title = "Party Composition by Region",
    x = "Region",
    y = "Count",
    fill = "Party"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```

Northeast and West regions are more Democratic than Republican and Midwest and South regions are more Republican than Democratic.

```{r}

# Region and Sentiment
sentiment_region_props <- df %>%
  group_by(region, model_label) %>%
  summarise(Count = n(), .groups = "drop") %>%
  group_by(region) %>%
  mutate(Proportion = Count / sum(Count))

ggplot(sentiment_region_props, aes(x = region, y = Proportion, fill = as.factor(model_label))) +
  geom_bar(stat = "identity", position = "fill") +
  scale_fill_manual(values = c("-1" = "red", "1" = "#078d0f"), labels = c("Risk", "Opportunity")) +
  labs(
    title = "Sentiment Distribution by Region",
    x = "Region",
    y = "Proportion",
    fill = "Sentiment"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```

Northeast and West regions are more opportunity than risk and Midwest and South regions are more risk than opportunity.

```{r}

# Boxplot for ideology score by region
ggplot(df, aes(x = region, y = ideology_score, fill = region)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_brewer(palette = "Set3") +
  labs(
    title = "Ideology Score Distribution by Region",
    x = "Region",
    y = "Ideology Score (-1 = Liberal, 1 = Conservative)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )

```

The Northeast is predominantly liberal with minimal variation, while the West and Midwest exhibit the greatest ideological diversity, spanning from strongly liberal to strongly conservative. The South leans slightly conservative but is less polarized compared to other regions.

```{r}

heatmap_data <- df %>%
  mutate(
    model_label = ifelse(model_label == 1, "Opportunity", "Risk")  
  ) %>%
  group_by(region, model_label) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(region) %>%
  mutate(percentage = (count / sum(count)) * 100) %>%
  ungroup()

# Heatmap
ggplot(heatmap_data, aes(x = model_label, y = region, fill = percentage)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.1f%%", percentage)), 
            color = "black", size = 4) +
  scale_fill_gradientn(
    colors = c("darkred", "red", "yellow", "lightgreen", "darkgreen"),  
    values = scales::rescale(c(0, 25, 50, 75, 100)), 
    name = "Percentage"
  ) +
  labs(
    title = "Heatmap of Sentiment Distribution by Region",
    x = "Sentiment",
    y = "Region"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    axis.text.y = element_text(hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )

```

The Northeast is the most optimistic region, strongly favoring opportunity. The Midwest is the most divided, highlighting significant differences in sentiment within the region. The West and South share similar sentiment distributions, leaning optimistic but still spreading risk regarding climate change.

```{r}

flows_region <- df %>%
  mutate(
    sentiment_label = ifelse(model_label == 1, "Opportunity", "Risk")  
  ) %>%
  group_by(region, sentiment_label) %>%
  summarise(value = n(), .groups = "drop") 

nodes_region <- data.frame(
  name = c(
    unique(flows_region$region),     
    unique(flows_region$sentiment_label)  
  )
)

links_region <- flows_region %>%
  mutate(
    source = match(region, nodes_region$name) - 1,       
    target = match(sentiment_label, nodes_region$name) - 1,  
    value = value                                   
  )

sankeyNetwork(
  Links = links_region, 
  Nodes = nodes_region,
  Source = "source",
  Target = "target",
  Value = "value", 
  NodeID = "name",
  sinksRight = TRUE, 
  nodeWidth = 30,
  fontSize = 12, 
  height = 600,
  width = 1000
)

```

```{r}

flows_region <- df %>%
  mutate(
    sentiment_label = ifelse(model_label == 1, "Opportunity", "Risk")  
  ) %>%
  group_by(region, sentiment_label) %>%
  summarise(value = n(), .groups = "drop") 

nodes_region <- data.frame(
  name = c(
    unique(flows_region$region),     
    unique(flows_region$sentiment_label)  
  )
)

links_region <- flows_region %>%
  mutate(
    source = match(region, nodes_region$name) - 1,       
    target = match(sentiment_label, nodes_region$name) - 1,  
    value = value                                   
  )

sankeyNetwork(
  Links = links_region, 
  Nodes = nodes_region,
  Source = "source",
  Target = "target",
  Value = "value", 
  NodeID = "name",
  sinksRight = TRUE, 
  nodeWidth = 30,
  fontSize = 12, 
  height = 600,
  width = 1000
)

```

To confirm this relationship between region and sentiment, we will run a correlation analysis.

```{r}

# Correlation test between ideology score and sentiment
ideology_correlation <- cor.test(df$ideology_score, df$model_label)
print("\nCorrelation between Ideology Score and Sentiment:")
print(ideology_correlation)

```

The correlation coefficient of -0.813 indicates a strong negative correlation between these two variables. This means that as a senator's ideology score increases, their sentiment tends to shift towards perceiving climate change in a risky way rather than an opportunistic way.

## Conclusion

The analysis reveals a divide in how U.S. senators view climate change. Democrats and Independents tend to see it in an opportunistic way, while Republicans are more likely to view it in a risky way. This divide is reflected in the negative correlation between a senator's ideology and their climate sentiment. The regional analysis shows the Northeast and West are more opportunity-focused, while the Midwest and South have a mix of risky and opportunistic perspectives. Some states have a more balanced, "mixed" sentiment, suggesting potential for bipartisan cooperation.
